Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-11-28 proposal/proposal.tex
Added Buy/Quote Client-Server example to parallel C&P's.

@@

@@

\newcommand{\fpop}{System F${}^\circ$\xspace}
\newcommand{\1}{\textbf{1}\xspace}

@@

Additionally, the majority of previous session type systems for mainstream
programming languages have focussed exclusively on linear type systems. One
disadvantage of a linear type system is the need to re-bind objects after each
operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide
the benefits of aliasing with the guarantees of linearity which has been
studied in several systems using a variety of alias control
methods~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}. Another approach,
which has been utilised in previous session-based $\lambda$-calculus type
systems, is to support side-effecting functions by allowing functions to close
over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}. An
interesting extension to the base GV type system would be to incorporate the
closure semantics into the type system which would permit functions to modify
session-typed channels in the environment. The typing rules would be similar
to that described by \citeauthor{Gay:2010:MST} for typing expressions in an
object-oriented setting for session types~\cite{Gay:2010:MST}, e.g. the rule
for lambda abstraction could be:

%% \begin{prooftree}
%% \AxiomC{$\Gamma , x \colon A \triangleright t \colon T \triangleleft \Gamma ' , x \colon B$}
%% \UnaryInfC{$\Gamma \triangleright t \colon (A \gg B) \rightarrow T \triangleleft \Gamma '$} 
%% \end{prooftree}

\begin{gather*}
  \frac{
    \typeds{\Gamma,\typedid{x}{A}}{t}{T}{\Gamma',\typedid{x}{B}}
  }{
    \typeds{\Gamma}{\abs{x}{t}}{\fn{\typestate{A}{B}}{T}}{\Gamma'}
  }\quad
\end{gather*}

@@

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL:

\input{sill-branch-choice-rules}

Note that branch and choice both have some notion of branching (R rules for
$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
a session-based setting, since the intutition for \& is that of
\emph{offering} a choice between alternatives rather than \emph{making} the
choice. Here, however, the intuition should be extended to account for its
location in the typing judgement. The channel is on the left of the sequent,
and \citeauthor{Caires:2010:STI} suggest one consider the context as services
being provided to the typed process. Thus, the type of channels in the context
is really the type of the other endpoint of the channel, and the typed process
behaves according to the dual protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to represent the dual of a protocol $A$, the typed process is behaving according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from alternatives (in this case, $\inrterm$) and then behaving according to the selected protocol (in this case, protocol $B^\bot$).

While in the case of branch and choice, which have different semantics this is
mitigated somewhat but as can be seen, there are two rules for selecting a
branch, one for branch and another for choice if we consider their server
example

As noted by Wadler...jfp there is a certain amount of redundancy in the
two-sided nature of the rules forcing one to commit to a certain convention
when typing a communication system. Consider, an equivalent system to the
buy/quote example given by caires and pfen. In the following, the types Book,
Quote and Receipt can be considered synoms for the type \1 as in caires and
pfenning.

... server example here ...

The derivation for ClientBody is T1R, T implies R, T tensor R, T plus
R2. Except for the final rule, these are exact opposites of the rules used for
...

The final system is near identical to the one produced by the example given
caires and pfenning. Here the client communicates along the channel $c$ on the
right and the server utilises the channel $c$ on the left. This typing, while
correct, violates the intuition given by caires and pfenning of the context
providing the services used by the process. The example above suggests the
server utilises some channel provided by the client. On the other hand, one
could imagine such a channel being setup between two communicating agents; the
``owner'' of the channel is largely irrelevant. Note that clientproto is in
fact the dual of serverproto in Wadler's classical process calculus.

most important: duality of sessions is obscured in $\pi$DILL because of the
nature of the typing judgements

>>>>

% Thanks to Stefan Kottwitz on tex.stackexhange.com:
% http://tex.stackexchange.com/questions/69662/how-to-globally-change-the-spacing-around-equations
%% \expandafter\def\expandafter\normalsize\expandafter{%
%%     \normalsize
%%     \setlength\abovedisplayskip{0pt}
%%     \setlength\belowdisplayskip{0pt}
%%     \setlength\abovedisplayshortskip{0pt}
%%     \setlength\belowdisplayshortskip{0pt}
%% }

@@

\allowdisplaybreaks

@@

\newcommand{\fpop}{System F${}^\circ$\xspace}

@@

Additionally, the majority of previous session type systems for mainstream
programming languages have focussed exclusively on linear type systems. One
disadvantage of a linear type system is the need to re-bind objects after each
operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide
the benefits of aliasing with the guarantees of linearity which has been
studied in several systems using a variety of alias control
methods~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}. Another approach,
which has been utilised in previous session-based $\lambda$-calculus type
systems, is to support side-effecting functions by allowing functions to close
over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}. An
interesting extension to the base GV type system would be to incorporate the
closure semantics into the type system which would permit functions to modify
session-typed channels in the environment. The typing rules would be similar
to that described by \citeauthor{Gay:2010:MST} for typing expressions in an
object-oriented setting for session types~\cite{Gay:2010:MST}, e.g. the rule
for lambda abstraction could be:
\begin{gather*}
  \frac{
    \typeds{\Gamma,\typedid{x}{A}}{t}{T}{\Gamma',\typedid{x}{B}}
  }{
    \typeds{\Gamma}{\abs{x}{t}}{\fn{\typestate{A}{B}}{T}}{\Gamma'}
  }\quad
\end{gather*}

@@

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL:
\input{sill-branch-choice-rules}

Note that branch and choice both have some notion of branching (R rules for
$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
a session-based setting, since the intutition for \& is that of
\emph{offering} a choice between alternatives rather than \emph{making} the
choice. Here, however, the intuition should be extended to account for its
location in the typing judgement. The channel is on the left of the sequent,
and \citeauthor{Caires:2010:STI} suggest one consider the context as services
being provided to the typed process. Thus, the type of channels in the context
is really the type of the other endpoint of the channel, and the typed process
behaves according to the dual protocol. For example, considering
$\SILLbranchLTwo$, and taking $A^\bot$ to represent the dual of a protocol $A$
as defined by \citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process
is behaving according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting
from alternatives (in this case, $\inrterm$) and then behaving according to
the selected protocol (in this case, protocol $B^\bot$).

As noted by Wadler~\cite{Wadler:2014} there is a certain amount of redundancy
in the two-sided nature of the rules forcing one to commit to a certain
convention when typing a communication system. Consider, an equivalent system
to the buy/quote example given by
\citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
\begin{gather*}
\sdef{ClientProto}{\BChoice{(\tensor{Book}
                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
\end{gather*}

In the protocol defined above, a client selects between buying a book or
receiving a price quote for a book. In the buy option case, the client sends
the name of the book to the server and their card details, then the server
sends the receipt for the purchase. In the quote option case, the client sends
the name of the book and receives a price quote from the server. I have
replaced the types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more
meaningful type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate;
all are synonyms for the type $\1$ which denotes the terminal session
type. The client and server process bodies are as follows:
\begin{gather*}
\sdef{ClientBody_c}
     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
\sdef{ServerBody_c}
     {\CPcase{c}
             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
\end{gather*}

\begin{samepage}
The typings for the client and server processes are given by:
\begin{gather*}
\tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
\\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
\end{gather*}
\end{samepage}

The derivation for \eqref{eq:cbtyp} is $T\1 R$, $\lfn{T}{R}$, $\tensor{T}{R}$,
$\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
composite process:
\begin{gather*}
\sdef{QSimple}{\Spar{c}{ClientBody_c}{ServerBody_c}}\\
\tsequent{\cdot}{\cdot}{QSimple}{\typedid{-}{\1}}
\end{gather*}

The final system is near identical to the one produced in the example given by
\citeauthor{Caires:2010:STI} except the individual processes within the
parallel composition have been swapped. Here the client communicates along the
channel $c$ on the right-side of the sequent and the server utilises the
channel $c$ on the left of the sequent. This typing, while correct, violates
the intuition given by \citeauthor{Caires:2010:STI} of the context providing
the services used by the process, since one would usually associate provision
of services with the server not the client. Instead, the example above
suggests the server utilises some channel provided by the client. On the other
hand, one could imagine such a channel being setup between two communicating
agents; the ``owner'' of the channel is largely irrelevant. Note that
$ClientProto$ is in fact the dual of the $ServerProto$ definition by
\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
classical process calculus~\cite{Wadler:2014}. In $pi$DILL, the duality
between endpoints of the same channel are obscured by the nature of the typing
judgements.
